### 2016-08-04 capability-machine.tex
Added short proofs for a couple of the lemmas.

\newcommand{\FS}{\var{FS}}

@@

The ticket dispenser test program uses $\malloc$. To be able to reason about programs that use $\malloc$, we assume that it follows the following specification. If you provide it with one argument, namely the size of the allocation you wish and a return capability, then at some point $\pcreg$ will contain an executable capability based on this return capability and the return register will contain a $\rwx$ capability for the freshly allocated memory. 
\begin{specification}[Malloc v.2]
  \begin{align*}
    &\forall \Phi \in \Mems \ldotp \forall \hsfoot, \hsframe \in \HeapSegments \ldotp \\
    &\quad \forall n, \size \in \nats \ldotp \forall b,e,a,m_b,m_e \in \Addrs\ldotp\forall p \in \Perms \ldotp \forall W \in \Worlds \ldotp\\
    &\qquad \exists i, \iota_{\malloc} \ldotp \\
    &\qquad \quad W = [i \mapsto \iota_{\malloc}] \land \\
    &\qquad \quad \dom(\hsfoot) = [m_b,m_e] \land \\ % This
    &\qquad \quad \memheap = \hsfoot \uplus \hsframe \land\\
    &\qquad \quad \heapSat[\hsfoot]{n}{W} \land \\
    &\qquad \quad \memreg(r_1) = \size \land \\
    &\qquad \quad \size \geq 0 \land \\
    &\qquad \quad \memreg(r_0) = (p,b,e,a) \land \\
    &\qquad \quad \memreg(\pcreg) = (\entry,m_b,m_e,m_b) \\ %and this needs to change with ABI.
    &\qquad \quad \Rightarrow \\
    &\qquad \qquad\exists \Phi' \in \Mems \ldotp \exists \hsfoot' \in \HeapSegments\ldotp\\
    &\qquad \qquad \quad \exists j,k \in \nats \ldotp \exists b',e'\in \Addrs \ldotp \exists W' \in \Worlds \ldotp \\
    &\qquad \qquad \qquad \Phi \step[j]\Phi' \land \\
    &\qquad \qquad \qquad \memheap[\Phi']=\hsfoot' \uplus \hsframe \land\\
    &\qquad \qquad \qquad W' = [k \mapsto \iota_{b',e'}^0, i \mapsto \iota_{\malloc}]\\
    &\qquad \qquad \qquad W' \future W \land \\
    &\qquad \qquad \qquad [b',e'] \subseteq \FS  \land \\
    &\qquad \qquad \qquad \heapSat[\hsfoot']{n-j}{W'} \land \\
    &\qquad \qquad \qquad \memreg[\Phi'](\pcreg) = (\updatePcPerm{p},b,e,a) \land \\
    &\qquad \qquad \qquad \size = e'-b' \\
    &\qquad \qquad \qquad \memreg[\Phi'](r_1) = (\rwx,b',e',b') \land \\
    &\qquad \qquad \qquad \forall r \in \RegName \setminus \{\pcreg,r_1\} \ldotp \memreg[\Phi'](r) = 0
\end{align*}
\end{specification}
The specification uses the mallog island defined as
\lau{I am not satisfied with the pc in the premise. The exact capability necessary very much depent on the implementation (this is mostly a note to myself.)}
\dominique{I agree with the remark about the pc.  Perhaps leave the form of the
  pc unspecified as it is not important for malloc users?}
\dominique{Resulting frame should be equal to original.}
\lau{Rewrite the specification when we have settled on a reasonable way to handle ABI.}

@@

\begin{align*}
  \Regions = \{ r = (s, \phi, H) \mid &  s \in \States \land \phi \subseteq States^2 \land \\ 
                                      &  H : \States \rightarrow \StorePred \land \\ 
                                      &  \phi \text{ reflexive and transitive} \}
\end{align*}
For convenience, we write $r.s$, $s.\phi$, and $s.H$ for a the projection of a regions state, transition system, and store predicate, respectively
\[
  \StorePred = \widehat{\Worlds} \monnefun \UPred{\HeapSegments}
\]

\[
\roll : \frac{1}{2} \cdot \Worlds \cong \widehat{\Worlds}
\]
Future Worlds:
\[
 W' \future W \text{ iff }  
 \begin{gathered}
   \dom(W') \supseteq \dom(W) \land \\
   \forall r \in dom(W) \ldotp W'(r) \future W(r)
 \end{gathered}
\]
Future regions:
\[
  (s_2,\phi_2,H_2) \future (s_1,\phi_1,H_1) \text{ iff } 
    (\phi_2,H_2) = (\phi_1,H_1) \land (s_1,s_2) \in \phi_1
\]
\begin{definition}[Heap satisfaction/erasure]
\[
\heapSat[\hs]{n}{W} \text{ iff }
\exists R \ldotp 
\begin{gathered}
  \dom(W) \rightarrow \HeapSegments \ldotp hs = \biguplus_{r \in \dom(W)} R(r) \\
  \text{and}\\
  \forall r \in \dom(W) \ldotp \forall n' < n \ldotp \npair[n']{R(r)} \in W(r).H(W(r).s)(\roll\; W)
\end{gathered}
\]
\end{definition}

@@

Consider what the ``good observations'' should be? Before we do that - what should the end result of a computation be?
\begin{align*}
  \observations \quad : & \quad  \Worlds \monfun \UPred{\Regs \times \HeapSegments} \\
  \observations (W) \defeq & \{ \npair{(\reg,\hs)} \mid \\
                           & \quad (\forall \heap_f, \heap', i \leq n \ldotp (\reg,\hs \uplus \heap_f) \step[i] (\halted,\heap')  \\
                           & \qquad \Rightarrow \exists W' \future W \ldotp\exists \hs' \ldotp \heap' = \hs' \uplus \heap_f \land \heapSat[\hs']{n-i}{W'}
\end{align*}
Removed the condition $\npair[n-i]{\heap'(0)} \in \stdvr(W)$.\\

@@

``Expression'' relation:
\begin{align*}
  \stder \quad : & \quad \Worlds \monfun \UPred{\Words} \\%TODO: Check whether this needs to be monotone
  \stder(W) \defeq & \{ \npair{\pc} \mid \\
                   & \quad \forall \npair{\reg} \in \stdrr(W) \ldotp \\
                   & \qquad \forall \npair{c} \in \stdkr(W) \ldotp \\
                   & \qquad \quad  \forall \heapSat[\hs]{n}{W} \ldotp \\
                   & \qquad \qquad \npair{(\reg\update{r_0}{c}\update{\pcreg}{\pc},\hs)} \in \observations(W) \}
\end{align*}

@@

\begin{lemma}[Heap satisfaction uniformity]
\[
  \forall n,n',\heap, W \ldotp \heapSat{n}{W} \land n' < n \Rightarrow \heapSat{n'}{W}
\]
\end{lemma}

\begin{lemma}[Read condition uniformity]
\[
\forall n,n',W,\start, \addrend \ldotp \readCond{n,W,\start,\addrend} \land n' \leq n \Rightarrow \readCond{n',W,\start,\addrend}
\]
\end{lemma}
\begin{proof}

\end{proof}

\begin{lemma}[Read-write condition uniformity]
\[
\forall n,n',W,\start, \addrend \ldotp \writeCond{n,W,\start,\addrend} \land n' \leq n \Rightarrow \writeCond{n',W,\start,\addrend}
\]
\end{lemma}

\begin{lemma}[Execute condition uniformity]
\[
\forall n,n',W,\start, \addrend, \perm \ldotp \execCond{n,W,\start,\addrend,\perm} \land n' \leq n \Rightarrow \execCond{n',W,\start,\addrend,\perm}
\]
\end{lemma}

\begin{lemma}[Entry condition uniformity]
\[
\forall n,n',W,\start, \addrend, \addr \ldotp \entryCond{n,W,\start,\addrend} \land n' \leq n \Rightarrow \entryCond{n',W,\start,\addrend,\addr}
\]
\end{lemma}

\begin{lemma}[Write condition implies read condition]
\[
\forall n, W, \start, \addrend \ldotp \writeCond{n,W,\start,\addrend} \Rightarrow \readCond{n,W,\start,\addrend}
\]
\end{lemma}

\begin{lemma}
  \[
    \forall n' < n \ldotp \forall W \ldotp \forall w \ldotp \npair{w} \in \stdvr(W) \Rightarrow \npair[n']{w} \in \stdvr(W)
  \]
\end{lemma}
\begin{proof}
  Follows from the uniformity of $\readCond{}$, $\writeCond{}$, $\execCond{}$, and $\entryCond{}$.
\end{proof}

\begin{lemma}
  \[
    \forall n \ldotp \forall W' \future W \ldotp \forall w \ldotp \npair{w} \in \stdvr(W) \Rightarrow \npair{w} \in \stdvr(W')
  \]
\end{lemma}

\begin{lemma}
  \[
    \forall n' < n \ldotp \forall W \ldotp \forall \reg \ldotp \forall \hs \ldotp \npair{(\reg,\hs)} \in \observations(W) \Rightarrow \npair[n']{(\reg,\hs)} \in \observations(W)
  \]
\end{lemma}

\begin{lemma}
  \[
    \forall n \ldotp \forall W \future W' \ldotp \forall \reg \ldotp \forall \hs \ldotp \npair{(\reg,\hs)} \in \observations(W) \Rightarrow \npair{(\reg,\hs)}(W')
  \]
\end{lemma}

\begin{lemma}
  \[
    \forall n' < n \ldotp \forall W \ldotp \forall \reg \ldotp \npair{\reg} \in \stdrr(W) \Rightarrow \npair[n']{\reg} \in \stdrr(W)
  \]
\end{lemma}

\begin{lemma}
  \[
    \forall n \ldotp \forall W' \future W \ldotp \forall \reg \ldotp \npair{\reg} \in \stdrr(W) \Rightarrow \npair{\reg} \in \stdrr(W')
  \]
\end{lemma}

\begin{lemma}
  \[
    \forall n' < n \ldotp \forall W \ldotp \forall c \ldotp \npair{c} \in \stdkr(W) \Rightarrow \npair[n']{c} \in \stdkr(W)
  \]
\end{lemma}

\begin{lemma}
  \[
    \forall n \ldotp \forall W' \future W \ldotp \forall c \ldotp \npair{c} \in \stdkr(W) \Rightarrow \npair{c} \in \stdkr(W')
  \]
\end{lemma}

\begin{lemma}
  \[
    \forall n' < n \ldotp \forall W \ldotp \forall \pc \ldotp \npair{\pc} \in \stder(W) \Rightarrow \npair[n']{\pc} \in \stder(W)
  \]
\end{lemma}

\begin{lemma}
  \[
    \forall n \ldotp \forall W' \future W \ldotp \forall \pc \ldotp \npair{c} \in \stder(W) \Rightarrow \npair{c} \in \stder(W')
  \]
\end{lemma}

>>>>

\newcommand{\base}{\var{base}}
\newcommand{\eend}{\var{end}}

@@

The ticket dispenser test program uses $\malloc$. To be able to reason about programs that use $\malloc$, we assume that it follows the following specification. If you provide it with one argument, namely the size of the allocation you wish and a return capability, then at some point $\pcreg$ will contain an executable capability based on this return capability and the return register will contain a $\rwx$ capability for the freshly allocated memory. 
\begin{specification}[Malloc v.2]
  \begin{align*}
    &\forall \Phi \in \Mems \ldotp \forall \hsfoot, \hsframe \in \HeapSegments \ldotp \\
    &\quad \forall n, \size \in \nats \ldotp \forall b,e,a,m_b,m_e \in \Addrs\ldotp\forall p \in \Perms \ldotp \forall W \in \Worlds \ldotp\\
    &\qquad \exists i, \iota_{\malloc} \ldotp \\
    &\qquad \quad W = [i \mapsto \iota_{\malloc}] \land \\
    &\qquad \quad \dom(\hsfoot) = [m_b,m_e] \land \\ % This
    &\qquad \quad \memheap = \hsfoot \uplus \hsframe \land\\
    &\qquad \quad \heapSat[\hsfoot]{n}{W} \land \\
    &\qquad \quad \memreg(r_1) = \size \land \\
    &\qquad \quad \size \geq 0 \land \\
    &\qquad \quad \memreg(r_0) = (p,b,e,a) \land \\
    &\qquad \quad \memreg(\pcreg) = (\entry,m_b,m_e,m_b) \\ %and this needs to change with ABI.
    &\qquad \quad \Rightarrow \\
    &\qquad \qquad\exists \Phi' \in \Mems \ldotp \exists \hsfoot' \in \HeapSegments\ldotp\\
    &\qquad \qquad \quad \exists j,k \in \nats \ldotp \exists b',e'\in \Addrs \ldotp \exists W' \in \Worlds \ldotp \\
    &\qquad \qquad \qquad \Phi \step[j]\Phi' \land \\
    &\qquad \qquad \qquad \memheap[\Phi']=\hsfoot' \uplus \hsframe \land\\
    &\qquad \qquad \qquad W' = [k \mapsto \iota_{b',e'}^0, i \mapsto \iota_{\malloc}]\\
    &\qquad \qquad \qquad W' \future W \land \\
    &\qquad \qquad \qquad \heapSat[\hsfoot']{n-j}{W'} \land \\
    &\qquad \qquad \qquad \memreg[\Phi'](\pcreg) = (\updatePcPerm{p},b,e,a) \land \\
    &\qquad \qquad \qquad \size = e'-b' \\
    &\qquad \qquad \qquad \memreg[\Phi'](r_1) = (\rwx,b',e',b') \land \\
    &\qquad \qquad \qquad \forall r \in \RegName \setminus \{\pcreg,r_1\} \ldotp \memreg[\Phi'](r) = 0
  \end{align*}
\end{specification}
The specification uses the mallog island defined as
\lau{I am not satisfied with the pc in the premise. The exact capability necessary very much depent on the implementation (this is mostly a note to myself.)}
\dominique{I agree with the remark about the pc.  Perhaps leave the form of the
  pc unspecified as it is not important for malloc users?}
\dominique{Resulting frame should be equal to original.}
\lau{Rewrite the specification when we have settled on a reasonable way to handle ABI.}

@@

\begin{align*}
  \Regions = \{ r = (s, \phi, H) \mid &  s \in \States \land \phi \subseteq \States^2 \land \\ 
                                      &  H : \States \rightarrow \StorePred \land \\ 
                                      &  \phi \text{ reflexive and transitive} \}
\end{align*}
For convenience, we write $r.s$, $s.\phi$, and $s.H$ for a the projection of a regions state, transition system, and store predicate, respectively
\[
  \StorePred = \widehat{\Worlds} \monnefun \UPred{\HeapSegments}
\]

\[
\roll : \frac{1}{2} \cdot \Worlds \cong \widehat{\Worlds}
\]
Future Worlds:
\[
 W' \future W \text{ iff }  
 \begin{gathered}
   \dom(W') \supseteq \dom(W) \land \\
   \forall r \in dom(W) \ldotp W'(r) \future W(r)
 \end{gathered}
\]
Future regions:
\[
  (s_2,\phi_2,H_2) \future (s_1,\phi_1,H_1) \text{ iff } 
    (\phi_2,H_2) = (\phi_1,H_1) \land (s_1,s_2) \in \phi_1
\]
\begin{definition}[Heap satisfaction/erasure]
\[
\heapSat[\hs]{n}{W} \text{ iff }
\begin{aligned}
&\exists R : \dom(W) \rightarrow \HeapSegments \ldotp \\
&\quad\begin{gathered}
   hs = \biguplus_{r \in \dom(W)} R(r) \\
  \text{and}\\
  \forall r \in \dom(W) \ldotp \forall n' < n \ldotp \npair[n']{R(r)} \in W(r).H(W(r).s)(\roll\; W)
\end{gathered}
\end{aligned}
\]
\end{definition}

@@

Consider what the ``good observations'' should be? Before we do that - what should the end result of a computation be?
\begin{align*}
  \observations \quad : & \quad  \Worlds \rightarrow \UPred{\Regs \times \HeapSegments} \\
  \observations (W) \defeq & \{ \npair{(\reg,\hs)} \mid \\
                           & \quad (\forall \heap_f, \heap', i \leq n \ldotp (\reg,\hs \uplus \heap_f) \step[i] (\halted,\heap')  \\
                           & \qquad \Rightarrow \exists W' \future W \ldotp\exists \hs' \ldotp \heap' = \hs' \uplus \heap_f \land \heapSat[\hs']{n-i}{W'}
\end{align*}
Removed the condition $\npair[n-i]{\heap'(0)} \in \stdvr(W)$.\\

@@

``Expression'' relation:
\begin{align*}
  \stder \quad : & \quad \Worlds \rightarrow \UPred{\Words} \\%TODO: Check whether this needs to be monotone
  \stder(W) \defeq & \{ \npair{\pc} \mid \\
                   & \quad \forall \npair{\reg} \in \stdrr(W) \ldotp \\
                   & \qquad \forall \npair{c} \in \stdkr(W) \ldotp \\
                   & \qquad \quad  \forall \heapSat[\hs]{n}{W} \ldotp \\
                   & \qquad \qquad \npair{(\reg\update{r_0}{c}\update{\pcreg}{\pc},\hs)} \in \observations(W) \}
\end{align*}

@@

\begin{lemma}[Heap satisfaction uniformity]
  \begin{align*}
    & \forall n,n',\heap, W \ldotp\\
    & \quad \heapSat{n}{W} \land n' < n \Rightarrow \heapSat{n'}{W}
  \end{align*}
\end{lemma}

\begin{lemma}[Read condition uniformity]
  \begin{align*}
    &\forall n,n',W,\start, \addrend \ldotp \\
    &\quad  \readCond{n,W,\start,\addrend} \land \\
    &\quad  n' \leq n \\
    &\qquad \Rightarrow \readCond{n',W,\start,\addrend}
  \end{align*}
\end{lemma}
\begin{proof}
  Follows directly from definition: Let $n$, $n'$, $\start$, and  $\addrend$ be given. Assume $\readCond{n,W,\start,\addrend}$ and $n' \leq n$. If $n=n'$, then the result follows trivially, as it is exactly our assumption. If $n' < n$, then by assumption we have a region name $r$, interval $[\start',\addrend'] \supseteq [\start,\addrend]$, and the assumption $W(r) \nsubeq[n-1] \iota_{\start',\addrend'}$  given by the read condition assumption. Picking $r$ and $[\start',\addrend']$, we need to show  $W(r) \nsubeq[n'-1] \iota_{\start',\addrend'}$ which follows from downwards closure of $\nsubeq$.
\end{proof}

\begin{lemma}[Read-write condition uniformity]
  \begin{align*}
    & \forall n,n',W,\start, \addrend \ldotp\\
    &\quad \writeCond{n,W,\start,\addrend} \land\\
    &\quad  n' \leq n\\
    &\qquad \Rightarrow \writeCond{n',W,\start,\addrend}
  \end{align*}
\end{lemma}
\begin{proof}
  Follows directly from the definition.
\end{proof}

\begin{lemma}[Execute condition uniformity]
  \begin{align*}
    &\forall n,n',W,\start, \addrend, \perm \ldotp \\
    &\quad  \execCond{n,W,\start,\addrend,\perm} \land \\
    &\quad  n' \leq n \\
    &\qquad \Rightarrow \execCond{n',W,\start,\addrend,\perm}
  \end{align*}
\end{lemma}
\begin{proof}
  Follows directly from definition: Let $n$, $n'$, $W$, $\start$, and $\addrend$ be given. Assume $\execCond{n,W,\start,\addrend,\perm}$ and $n' \leq n$. If $n' = n$, then the result follows directly from assumption. If $n' < n$, then let $n'' < n'$, $W' \future W$, and $a \in [\start,\addrend]$ be given. Using that we by transitivity have $n'' < n$, we get the desired result from assumption.
\end{proof}

\begin{lemma}[Entry condition uniformity]
  \begin{align*}
    & \forall n,n',W,\start, \addrend, \addr \ldotp\\
    &\quad  \entryCond{n,W,\start,\addrend,\addr} \land \\
    &\quad  n' \leq n \\
    &\qquad \Rightarrow \entryCond{n',W,\start,\addrend,\addr}
  \end{align*}
\end{lemma}
\begin{proof}
  Like the $\execCond{}$ uniformity proof.
\end{proof}

\begin{lemma}[Write condition implies read condition]
  \begin{align*}
    &\forall n, W, \start, \addrend \ldotp\\ 
    &\quad \writeCond{n,W,\start,\addrend} \Rightarrow \readCond{n,W,\start,\addrend}
  \end{align*}
\end{lemma}
\begin{proof}
  Follows directly from the definition.
\end{proof}

\begin{lemma}
  \begin{align*}
    &\forall n' < n \ldotp \forall W \ldotp \forall w \ldotp\\
    &\quad \npair{w} \in \stdvr(W) \Rightarrow \npair[n']{w} \in \stdvr(W)
  \end{align*}
\end{lemma}
\begin{proof}
  Follows from the uniformity of $\readCond{}$, $\writeCond{}$, $\execCond{}$, and $\entryCond{}$.
\end{proof}

\begin{lemma}
  \begin{align*}
    &\forall n \ldotp \forall W' \future W \ldotp \forall w \ldotp\\ 
    &\quad \npair{w} \in \stdvr(W) \Rightarrow \npair{w} \in \stdvr(W')
  \end{align*}
\end{lemma}

\begin{lemma}
  \begin{align*}
    &\forall n' < n \ldotp \forall W \ldotp \forall \reg \ldotp \forall \hs \ldotp\\
    &\quad \npair{(\reg,\hs)} \in \observations(W) \Rightarrow \npair[n']{(\reg,\hs)} \in \observations(W)
  \end{align*}
\end{lemma}

\begin{lemma}
  \begin{align*}
    &\forall n' < n \ldotp \forall W \ldotp \forall \reg \ldotp\\
    &\quad \npair{\reg} \in \stdrr(W) \Rightarrow \npair[n']{\reg} \in \stdrr(W)
  \end{align*}
\end{lemma}

\begin{lemma}
  \begin{align*}
    &\forall n \ldotp \forall W' \future W \ldotp \forall \reg \ldotp\\
    &\quad \npair{\reg} \in \stdrr(W) \Rightarrow \npair{\reg} \in \stdrr(W')
  \end{align*}
\end{lemma}

\begin{lemma}
  \begin{align*}
    &\forall n' < n \ldotp \forall W \ldotp \forall c \ldotp\\
    &\quad \npair{c} \in \stdkr(W) \Rightarrow \npair[n']{c} \in \stdkr(W)
  \end{align*}
\end{lemma}

\begin{lemma}
  \begin{align*}
    &\forall n \ldotp \forall W' \future W \ldotp \forall c \ldotp\\
    &\quad \npair{c} \in \stdkr(W) \Rightarrow \npair{c} \in \stdkr(W')
  \end{align*}
\end{lemma}

\begin{lemma}
  \begin{align*}
    & \forall n' < n \ldotp \forall W \ldotp \forall \pc \ldotp\\
    & \quad \npair{\pc} \in \stder(W) \Rightarrow \npair[n']{\pc} \in \stder(W)
  \end{align*}
\end{lemma}
